Parametricity Consequences of GADTs Representations

### 2021-02-15 20-gadts/GADT-story.tex
work on gadt-story Sec 2.2

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Eq}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $Eq$. Such a type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Eq :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Eq\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Eq\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Eq}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
         {\color{red}
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Eq\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]}
in a higher-kinded calculus such as $F_\omega$.

Importantly, $Eq\,a$ cannot be made into a functor, or, equivalently,
$\mathsf{Eq\,a}$ cannot be made an instance of Haskell's
$\mathsf{Functor}$ class: if $\mathsf{Eq\,a}$ supported a function
$\mathsf{map_{Eq\,a} :: (b \to c) \to Eq\,a\,b \to Eq\,a\,c}$ then
defining
\[\begin{array}{l}
\mathsf{eqElim :: Eq\, a\, b \to b \to a}\,{\color{red} was Eq\,a\,b\to a\to b}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the element 
\[\begin{array}{l}
\mathsf{emptyFromMapEqa :: (\forall b\, c.\, (b \to c) \to Eq\, a\,
  b \to Eq\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEqa \,map_{Eq\,a} = eqElim\, (map_{\Eq\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$, which is impossible. It follows immediately
from this contradiction that $\mathsf{Seq}$ cannot support a map
function either. This is already intuitively evident from the Haskell
definition of $\mathsf{Seq}$, though. Indeed, the clause of
$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

@@

The primitive representations of {\color{blue} polynomial} GADTs
capture in syntax the fact that GADTs can be considered as fixpoints
of higher-order functors. Syntactically, the use of higher-order
funct{\em ions} is essential, since the type arguments to the GADT
being defined are not uniform across all of its instances in the types
of its data constructors. Semantically, the use of higher-order
funct{\em ors} is essential, as for ADTs, to the very existence of the
fixpoints being computed~\cite{tfca}.
%We will consider GADTs'
%semantics in more detail in Section~\ref{sec:GADT-semantics}.

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
a b.\,f\,a \times f\,b\] 
where $\Lan_K \,G$ is the 
%$Lan_k\, g$ is the syntactic reflection of the
{\em left Kan extension}~\cite{mac71}
%$\Lan_K \,G$
of the functor $G$
%interpreting $\mathsf{g}$
along the functor $K$.
%interpreting $\mathsf{k}$~\cite{mac71}.
The left Kan extension $\Lan_K \,G$ is the smallest functor
approximating/extending $K$ and coinciding with $G$ {\color{blue}
  better explanation}.

LKE gives the functorial completion of the GADT.

%written as
%%\[\mathsf{Seq\,a} = \mathsf{a + Eq\,a\,(b \times c) \times Seq\,b
%%  \times Seq\,c} \] 
%\[\begin{array}{l}
%\mathsf{data\, Seq :: * \to *\,where}\\
%\mathsf{\;\;\;\;\;\;\;\;Const ::\, a \to Seq\,a}\\
%\mathsf{\;\;\;\;\;\;\;\;Pair\,\,\,\,\, ::\, Eq\,a\,(b \times c) \to
%  Seq \,b \to Seq\,c \to Seq\,a}\\ 
%\end{array}\]
%
%The above rewriting of $\mathsf{Seq}$'s definition shows that it can
%be seen as a fixpoint $\mu H$ of the higher-order function $H$ defined
%on objects by $H\,F\,a = a + Eq\,a\,(b \times c) \times F\,b \times
%F\,c$.

{\color{blue} Need syntax that reflects semantic Lan.}

%Other GADTs are similarly characterizable by higher-order functors.
%{\color{blue} GADTs make essential use of higher-order functors.}

Need syntax functors (Lan)
The representation choices are similar for GADTs. The difference is
that the two are not isomorphic.

Certain polynomial GADTs correspond to polynomial ADTs. So we expect
the same behaviors.  Polynomial nested types are actually the GADTs
that mimic polynomial ADT behavior.

But pattern matching fails for the extension (functorial completion)
of a GADT.

>>>>

To determine the possible shapes of a GADT's container we have to
pattern-match on the type of the data to be contained. For this, our
GADT calculus must support an equality type $\mathsf{Equal}$ that is
inhabited when its two type arguments are the same and is uninhabited
otherwise, i.e., is the syntactic reflection of semantic equality
function $\mathit{Equal}$. Such a type can either be taken as primitive, as in
\[\begin{array}{l}
\mathsf{data\, Equal :: * \to * \to *\,where}\\
\mathsf{\;\;\;\;\;\;\;\;Refl ::\, Equal\,a\,a}\\
\end{array}\]
\noindent
or
represented by its own Church encoding
\[\mathsf{Equal\,t_1\,t_2} = \mathsf{\forall f.\, (\forall a.\,
  f\,a\,a) \to f\,t_1\,t_2}\] Using $\mathsf{Equal}$, the GADT
$\mathsf{Seq}$ can then be represented as its Church encoding
\[\mathsf{Seq\,a} = \mathsf{\forall f.\, (\forall a.\,a \to f\, a)
  \to (\forall a\,b\,c.\,Equal\,a\,(b\times c) \to f\,b \to f\,c \to
  f\,a) \to f\,a}\]
in a higher-kinded calculus such as $F_\omega$.

Importantly, $\mathit{Equal}\,a$ cannot be made into a functor, or,
equivalently, $\mathsf{Equal\,a}$ cannot be made an instance of
Haskell's $\mathsf{Functor}$ class: if $\mathsf{Equal\,a}$ supported a
function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct the element 
\[\begin{array}{l}
\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
  b \to Equal\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$, which is impossible. It follows immediately
from this contradiction that $\mathsf{Seq}$ cannot support a map
function either. This is already intuitively evident from the Haskell
definition of $\mathsf{Seq}$, though. Indeed, the clause of
$\mathsf{map}$ for the $\mathsf{Pair}$ constructor should feed
$\mathsf{map}$ a function $\mathsf{f :: (b \times c) \to e}$ and a
term of the form $\mathsf{Pair \,t_1\,t_2}$ for $\mathsf{x :: b}$ and
$\mathsf{y :: c}$, and produce a term $\mathsf{Pair\,t_1'\,t_2'}$ for
some appropriately typed terms $\mathsf{t_1'}$ and
$\mathsf{t_2'}$. But it is not clear how to achieve this since
$\mathsf{e}$ need not necessarily be a product type. And even if
$\mathsf{e}$ {\em were} known to be of the form $\mathsf{w \times z}$,
then we still wouldn't necessarily have a way to produce data of type
$\mathsf{w \times z}$ from only $\mathsf{f :: b \times c \to w \times
  z}$ and $\mathsf{x}$ and $\mathsf{y}$ unless we knew, e.g., that
$\mathsf{f}$ was a pair of functions.

@@

The primitive representations of {\color{blue} polynomial} GADTs
capture in syntax the fact that GADTs can be considered as fixpoints
of higher-order functors. Syntactically, the use of higher-order
funct{\em ions} is essential, since the type arguments to the GADT
being defined are not uniform across all of its instances in the types
of its data constructors. Semantically, the use of higher-order
funct{\em ors} is essential, as for ADTs, to the very existence of the
fixpoints being computed~\cite{tfca}.
%We will consider GADTs' semantics in more detail in
%Section~\ref{sec:GADT-semantics}.

Consider, again, the GADT $\mathsf{Seq}$ from the start of
Section~\ref{sec:gadts}. Because its type argument varies in the
instances of $\mathsf{Seq}$ appearing in the types of its data
constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be characterized by
any first-order fixpoint equation. It can, however, be viewed as a
solution to the higher-order fixpoint equation
\[H\,f\,a = a \,+\,Lan_{\lambda a b.\,a \times b}\, \lambda
a b.\,f a \times f b\] where $\Lan_K \,G$ is the left Kan extension of
$G$ along $K$.  The left Kan extension $\Lan_K \,G : {\cal E} \to
{\cal D}$ of the functor $G : {\cal C} \to {\cal D}$ along the functor
$K : {\cal C} \to {\cal E}$ is the best functorial approximation to
$G$ that factors through $K$. Here, ``best functorial approximation''
means that $\Lan_K \,G$ is the smallest functor that both extends the
image of $K$ to $\cal D$ and coincides with $G$ on $\cal C$, in the
sense that, for any other such functor $H$, there is a morphism of
functors (i.e., a natural transformation) from $\Lan_K \,G$ to
$H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $G : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $G$ along $K$} is a
functor $\mathit{Lan}_K\,G : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : G \to (\mathit{Lan}_K\,G) \circ K$
such that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : G \to H \circ K$, there exists a unique
natural transformation $\mu : \mathit{Lan}_K\,G \to H$ such that $(\mu
K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{G}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,G}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\mu}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types a calculus must support a
primitive fixpoint constructor $\mathsf{\mu}$ that is the syntactic
reflection of the higher-order fixpoint operator on the category of
functors on the cagegory interpretating types, as well as a primitive
type constructor $\mathsf{Lan}$ such that $\mathsf{Lan_k\,g}$ is the
syntactic reflection of the left Kan extension $\Lan_K \,G$ of the
functor $G$ interpreting $\mathsf{g}$ along the functor $K$
interpreting $\mathsf{k}$. {\color{blue} This was elided in ADT
  case. Why? Tie together better.}  The type constructor
$\mathsf{Lan}$ thus computes the smallest data type that is generated
by a GADT's {\color{blue} GADT or constructor?} syntax and also
supports a $\mathsf{map}$ function. That is, $\mathsf{Lan}$ computes
the functorial completion of the data type determined by a GADT's
syntax.

Note, however, that the functorial completion of a GADT need not be a
nested type. As mentioned in the introduction, only those instances of
the GADT reachable from the syntax of the GADT are in the functorial
completion.

The representation choices are the same for GADTs as for ADTs. The
difference is that, although these two representations are isomorphic
for ADTs, they are not isomorphic for GADTs.  We will see this in
Section~\ref{??} below.
